Labor Markets and Technological Change: Evidence from Electronic Health Records

### 2022-06-13 Paper/Paper.tex

How technological innovation affects labor markets is often controversial, in part because of the vast differences in technology and tasks present even within industries. Moreover, shocks to an existing labor market could have downstream effects on customers in that market. A prime instance of this is in health care, where technological innovation in health care may contribute to physician burnout, which has been shown to affect patient satisfaction (\cite{shanafelt2002burnout}), or technological change may encourage physician retirement, possibly limiting patient access to care and ultimately affecting patient health. The most  widespread and health care-altering technological innovation in recent years took place during the implementation of electronic health records. This huge innovation in the health care sector, where there is extensive margin for improvement, is an excellent setting to examine the relationship between innovation and labor market, and the downstream effects this may have.  In this paper, I investigate physician labor market changes as a result of a major technology shift in the U.S. health care system, the implementation of electronic health records.

@@

Physicians, as primary users of this technology, play an important role in whether potential benefits of EHRs are realized. With the rapid implementation of a complex technology, the practical tasks that take place in a physician's daily life changed just as rapidly. In interviews, physicians reported that when using EHRs they are less satisfied with their job and have higher stress levels. Senior physicians in particular “loathe the cumbersome, time-consuming data entry that comes with using EHRs.” (\cite{CollierBurnout}). The frustration of using a new technology raises the cost of working in specific hospitals, which may lead physicians on the margin to make changes in labor market choices such as exiting the labor market altogether or shifting towards other work settings. However, the extent to which this frustration actually imposes a meaningful cost is unknown, making physician response an empirical question. Using a difference-in-differences research design, I estimate the effects of hospital EHR implementation on physician labor market outcomes.

My primary datasets are the CMS Shared Patient data and the Medicare Data on Provider Practice and Specialty (MD-PPAS), from which I construct a physician-level panel from 2009-2017 consisting of hospitalists and other general practice physicians who work exclusively in hospitals. The data captures whether a physician is exposed to EHRs in various hospitals, several labor market outcomes, and other demographic information. For all outcomes, the main independent variable of interest is a binary treatment variable capturing a physician's EHR exposure in hospitals. In the main analysis, a physician is considered to be exposed to an EHR if they are working in any hospital that implements an EHR. I estimate group time treatment effects of EHR exposure on the following physician decisions: (1) retirement, measured based on zero or missing billable activity in all future years of the panel, (2) where to physically work, measured by fraction of patients seen in an office setting, and list of primary zip codes, (3) number of patients seen, and (4) billing activity.

My analysis contributes to the literature examining effects of health information technology on health care. There is a robust group of empirical papers that examine the effects of EHRs on patient outcomes and hospital costs prior to the HITECH Act taking effect. Despite a large number of case studies that find generally positive effects (improved patient outcomes and decreased costs) (surveyed in \cite{Buntin2011TheResults}), more recent empirical work has found a mixture of results: that the median patient experiences no change due to EHRs (\cite{Agha2014TheCare}, \cite{McCullough2016HealthCoordination}, \cite{Meyerhoefer}) while newborns and severe patients experience improvement in health outcomes (\cite{Miller2009}, \cite{Freedman2015}, \cite{McCullough2016HealthCoordination}), and that hospital costs only decrease 6 years after implementation, if at all (\cite{Agha2014TheCare}, \cite{dranove2014trillion}). More relevant to EHR effects on physicians, a number of studies consider the effect of EHRs on productivity in specific settings. One study finds that nursing home productivity increases after adopting health IT (\cite{Hitt2016}), but another finds that physician productivity decreases by 11 percent due to EMRs being adopted in primary-care sites (\cite{Meyerhoefer}).

I expand and improve upon this literature, first by considering a different mechanism by which EHRs could indirectly affect patients through various physician responses, which, to my knowledge, has not yet been examined. This is the first empirical study that connects EHR implementation to physician labor market decisions, and focuses on the difference in response across age groups. Further, I improve on the health information technology literature by considering the primary time period in which EHRs were rapidly implemented. This provides sufficient variation in the timing of implementation and ensures the EHRs are advanced enough to be considered "meaningful" by the government's standards, and are likely affecting physicians' daily life. Similarly to past studies, I consider EHR implementation as a treatment variable in a difference-in-difference framework, but I improve on this by estimating average group time treatment effects for multiple years after implementation, which avoids common problems that arise in typical two way fixed effects estimation of heterogeneous effects.

Early research on the inputs to physician retirement decisions found that the most influential factors are personal and financial matters, and that physicians care about their work environment (specifically in the context of managed care organizations), but do not necessarily retire early because of it (\cite{Bahrami2002}). Contrary to this work, I find that EHR implementation led to a .005 ppt increase (25\% increase relative to the mean percentage) in the likelihood of a senior physician retiring in the year after implementation. Younger physicians (less than 60) are .002 ppts more likely to retire in the second year after exposure, possibly pointing to career switching. Whether physicians are retirement age or not, these results suggest that EHR implementation may have intensified access to care issues for patients, and that there was a surge of experienced physicians who left the labor force and were no longer available to influence physicians learning under them. In those who did not retire, I find that younger physicians are .015 ppts more likely to work in an office in the year of exposure to an EHR, but that exposure does not change the average fraction of patients seen in office settings for either age group. Finally, I find that both the total number of patients and billing activity increases in younger physicians in the first 2-3 years after exposure to an EHR. The ratio of billing activity to patient count increases less than the sample average, suggesting that billing activity did not dis-proportionally increase.

A key assumption that underlies the main analysis is that physicians are constrained to using the technology in a hospital who chooses to adopt. I consider two possibilities where this assumption is violated. First, I investigate the existence of hospital employees with the purpose of using an EHR on behalf of a physician, which I refer to as data assistants. If this is occurring, the results are at least partially driven by the hiring of new employees. I find that the majority of hiring of data assistants took place in 2013-2014, a two year delay from the majority of EHR implementation. Further, my main results are not sensitive to the exclusion of these years, and estimating the effect of EHR implementation only in hospitals without data assistants yields noisy, but similar, results. Second, I investigate whether the decision for hospitals to implement EHRs is exogenous to individual physicians. That is, a physician does not choose their own EHR exposure. While this assumption is reasonable in many settings, there may be instances where a physician is involved in hospital decision making. For this reason, I also consider a sensitivity analysis in which I limit the sample only to hospitals where vertical integration between physician and the hospital is low, as this indicates a lower probability that EHR implementation is endogenous. The results of the additional analyses indicate that my main findings are not sensitive to endogeneity concerns stemming from joint decision making among physicians and hospitals regarding EHR adoptions.

@@

This section serves to investigate why we might expect physicians to respond to the implementation of EHRs. The physician behaviors that I consider in this analysis are closely related and should not be taken as independent. For example, a physician's choice of work setting depends on their previous decision to prolong retirement, and billing activity depends on work setting. Thus, I discuss the behaviors in terms of stages of decision making.

@@

Workers in developed countries tend to plan for formal retirement well in advance. Generally, an exogenous shock to a worker is not expected to change retirement age simply because of the amount of wealth and planning necessary to formally leave the labor force. Classic retirement models structure the decision of retirement by maximizing expected lifetime utility based on future earnings and retirement benefits, or choosing to prolong retirement only if the expected earnings gain from doing so is positive (\cite{gustman1986disaggregated}, \cite{stock1990pension}). However, physicians typically make this decision differently than workers in other industries. Most physicians plan to retire at age 60, but do not actually retire until age 69 (\cite{collier2017challenges}). When the time to retire comes, many physicians hesitate to abandon patients they have seen over the course of their career. The decision to delay retirement is not financial, but altruistic. Thus, retiring may be in a physician's choice set years before the realized decision to leave the labor force.

It is important to note that age is the most important factor in the decision to retire, implying that a 35 year old physician has very different incentives than a 65 year old physician, and a technology shock will affect the age groups differently. First, I will walk through the incentives of a physician who is already of retirement age when the EHR is implemented. While payment schemes, income as a whole, and retirement benefits certainly impact the utility a physician receives from working or retiring, I can model the decision as independent of those things since I assume the physician already has the financial mean to retire. Then, a physician will only retire if the utility from working is less than the utility from retiring, assuming the physician keeps working under indifference. How does exogenous EHR implementation affect these utilities? The utility received in retirement will not be affected by whether the hospital is using an EHR after the physician is already gone. However, the utility received from working is a function of work environment. If EHRs impose a burden on daily tasks, there is an inverse relationship between EHR exposure and utility. If EHRs make daily tasks easier, there is a positive relationship. Thus, for a physician on the margin of retirement, the decision to retire only happens as a result of EHRs if they impose a cost, or a negative impact on work environment. Further, the cost has to be high enough to raise utility from retirement above utility from working, counteracting the altruism that is keeping the physician from retiring.

Formal retirement (leaving the labor force altogether) is not the only way for physicians to transition out of a clinical role. There are opportunities for physicians of any age to switch careers towards more administrative roles such as teaching, consulting, or hospital management. While it is unlikely that physicians under 55 are formally retiring, physicians of any age could be career switching. This decision is impacted similarly to the above discussion regarding formal retirement, but is also impacted by income that would be received as a result of career switching.

The term "retirement" for the purpose of this paper indicates no longer seeing patients, and makes no assertion about what the physician does afterwards. A physician is said to retire in a given year only if it is the first year in which the physician has no future Medicare claims.

As an exogenous shock to work environment, EHR implementation may lead to an earlier retirement than would have otherwise occurred. Media articles focusing on interviews with particular physicians suggest that the implementation of EHRs did in fact lead to retirement in older physicians (\cite{ringel_2019}, \cite{loria_2020}). The aim of this section is to test that hypothesis and evaluate whether younger physicians exhibited similar behavior.

@@

@@

The Medicare Data on Provider Practice and Specialty (MD-PPAS) is a database of physicians which details claim counts, physician specialty, and practice location from 2009-2017. Included in the MD-PPAS is physician NPI, primary and secondary specialties, both patient and total billing counts in up to 12 different zip codes, and fraction of patients in different settings such as office, inpatient or emergency room. I first limit the sample of physicians based on specialty type. My analysis relies on EHR implementation in hospitals. Thus, I only include physicians who reported their primary specialty as hospitalist (physician who self-identifies as hospitalist or has 90\% of patients in inpatient setting), internal medicine, pediatrics (although Medicare primarily enrolls the elderly, Medicare also covers disabled children), general practice or family practice in at least one year. However, I exclude physicians who list themselves as a specialist in all but one year. Since the focus of the analysis is physicians working in hospitals, I also drop physicians with less than 20\% of patients in a hospital setting in all years.

@@

I show summary statistics for the entire sample in Table \ref{tab:sumstats1}. Only 2\% of physicians (approximately 750) retire over the course of the panel. Just over half of the physician-year observations have value one for working in an office in some capacity, with about a third of the number of patients being seen in an office setting. By construction, all physicians are exposed to an EHR at some point in the sample, so the variation in treatment variables comes from the timing of exposure. Physicians are typically exposed about one fourth of the way through the sample, but there is a positive number of physicians first exposed in each year from 2009-2015. The average physician is 48 years old, works with 1.5 hospitals and 1.3 systems.

@@

In this setting, treatment effects are likely heterogeneous across physicians who are exposed in different years since the underlying characteristics of hospitals who adopt in different years may be correlated with the decision to adopt. Therefore, to avoid negative weighting issues this causes in a two way fixed effects specification, I estimate average treatment effects for a specific group $g$ at time $t$: 
$$ATT(g,t)=\mathbbm{E}[Y_t(g)-Y_t(0)|G_g=1],$$
where $G_g=1$ for those in group $g$. A group indicates all physicians treated, or first exposed to an EHR, in the same year. To estimate the heterogeneous treatment effects, I employ the estimator established in \citeauthor{callaway2021difference} (\citeyear{callaway2021difference}). Other estimators that similarly address the concerns yield similar results, presented in Appendix \ref{app:estimators}.

@@

For most physicians, EHRs will not impose a costly enough burden to induce retirement. However, there are other ways physicians might change their labor market behavior under sufficient incentive to avoid the technology. I consider the decision to switch work settings using multiple outcomes. First, I directly investigate the extent to which physicians work in office based setting and whether this changes due to EHR exposure. I use variation in two different variables: (1) an indicator variable equal to 1 if the physician has seen any positive number of patients in an office in a given year, and (2) the fraction of patients a physician sees in an office based setting in a given year. This analysis does not include physicians who retire at any point in the sample, as a zero could indicate dropping out of the data instead of working solely in a hospital.

@@

An important, and heavily debated, aspect of the implementation of EHRs is whether or not they allow users to be more productive. A main purpose of the HITECH Act was to improve efficiency of care, specifically to decrease time spent on administrative burden. Such a decrease implies a change in patient care: either a larger number of patients have access to an appointment, or the same amount of patients receive more face-to-face time with physicians during appointments. Alternatively, because the technology is not hugely user friendly, it may not decrease administrative burden after all. In this section, I speak to efficiency by estimating the effect of EHR implementation on patient count.

@@

\begin{figure}[p]
\centering
    \caption{Treatment Variables Over Time}
    \includegraphics[scale=.55]{Objects/sum_stats_year.pdf}
    \label{fig:treatmentgraph}
\end{figure}

@@

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Retirement}
    \includegraphics[scale=.45]{Objects/retire_plot.pdf}
    \label{fig:retirefirst}
\end{figure}

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Likelihood of Working in Office}
    \includegraphics[scale=.45]{Objects/officeind_plot.pdf}
    \label{fig:officefirst}
\end{figure}

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Fraction of Total Patients Seen in Office}
    \includegraphics[scale=.45]{Objects/officefrac_plot.pdf}
    \label{fig:officesecond}
\end{figure}

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Likelihood of Changing Zip Codes}
    \includegraphics[scale=.45]{Objects/zip_plot.pdf}
    \label{fig:zip}
\end{figure}

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Patient Count}
    \includegraphics[scale=.45]{Objects/patient_plot.pdf}
    \label{fig:patient}
\end{figure}

@@

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Claim Count, Limited Sample}
    \includegraphics[scale=.45]{Objects/claim_plot.pdf}
    \label{fig:claim}
\end{figure}

@@

I begin by connecting every National Provider Identifier (NPI)\footnote{Data can be downloaded from \hyperlink{https://download.cms.gov/nppes/NPI/Files.html}{\text{https://download.cms.gov/nppes/NPI-Files.html}}} to a description of their tax code\footnote{Data can be downloaded from \hyperlink{https://nucc.org/index.php/code-sets-mainmenu-41/provider-taxonomy-mainmenu-40/pdf-mainmenu-53}{\text{https://nucc.org/index.php/code-sets-mainmenu-41/provider-taxonomy-mainmenu-40/pdf-mainmenu-53}}}. I then categorize NPIs by key words in their tax code description. Any description containing ``Internal Medicine", ``Hospitalist", ``Family Medicine", ``General Practice", or ``Pediatrics" is classified as a primary care physician (PCP), and any description containing ``hospital" is classified as a hospital. I create two data sets, one containing only PCPs and one containing only hospitals. These data are saved to use in identifying the relevant NPIs in the CMS Shared Patient Data.

@@

@@

>>>>

How technological innovation affects labor markets is often controversial, in part because of the vast differences in technology and tasks present even within industries. Moreover, shocks to an existing labor market could have downstream effects on customers in that market. A prime instance of this is in health care, where technological innovation may affect patient care indirectly if it induces physician burnout, which has been shown to affect patient satisfaction (\cite{shanafelt2002burnout}), or if it brings about changes in the composition of aggregate physician labor markets in terms of total number of physicians, number of physicians working in rural settings, or number of physicians working in hospitals, which all effect patient access to care. The most widespread and health care-altering technological innovation in recent years took place with the implementation of electronic health records. Because the technology was incentivized by policy, it was implemented rapidly and on a large scale, making it an opportune setting to examine the relationship between labor markets and technology, and the potential downstream effects that took place. In this paper, I investigate physician labor market changes as a result of a major technology shift in the U.S. health care system, the implementation of electronic health records.

@@

Physicians, as primary users of this technology, play an important role in whether potential benefits of EHRs are realized. With the rapid implementation of a complex technology, the practical tasks that take place in a physician's daily life changed just as rapidly. In interviews, physicians reported that when using EHRs they are less satisfied with their job and have higher stress levels. Senior physicians in particular “loathe the cumbersome, time-consuming data entry that comes with using EHRs.” (\cite{CollierBurnout}). The frustration of using a new technology raises the cost of working in certain settings, which may lead physicians on the margin to make changes in labor market choices such as exiting the labor market altogether or shifting towards lower cost settings. However, the extent to which this frustration actually imposes a meaningful cost is unknown, making physician response an empirical question. Using an event study research design, I estimate the effect of hospital EHR implementation on individual physician labor market outcomes.

My primary datasets are the CMS Shared Patient data and the Medicare Data on Provider Practice and Specialty (MD-PPAS), from which I construct an individual-level panel from 2009-2017 consisting of physicians who work primarily in a hospital, known as hospitalists. The data captures whether such physicians are exposed to EHRs in their associated hospital(s), several labor market outcomes, and other demographic information. For all outcomes, the main independent variable of interest is a binary treatment variable capturing exposure to an EHR. In the main analysis, a physician is considered to be exposed to an EHR if they are working in any hospital that implements an EHR. I estimate group time treatment effects of EHR exposure on the following physician decisions: (1) retirement, measured based on zero or missing billable activity in all future years of the panel, (2) where to physically work, measured by fraction of patients seen in an office setting, and list of primary zip codes, (3) number of patients seen, and (4) billing activity.

This paper contributes to the literature's understanding of how health information technology generally affects health care. There is a robust set of empirical papers examining the effect of EHRs on patient outcomes and hospital costs, but these studies largely use data prior to the subsidization of meaningful EHR use. Despite a large number of case studies that find generally positive effects (improved patient outcomes and decreased costs) (surveyed in \cite{Buntin2011TheResults}), more recent empirical work has found a mixture of results: that the median patient experiences no change due to EHRs (\cite{Agha2014TheCare}, \cite{McCullough2016HealthCoordination}, \cite{Meyerhoefer}) while newborns and severe patients experience improvement in health outcomes (\cite{Miller2009}, \cite{Freedman2015}, \cite{McCullough2016HealthCoordination}), and that hospital costs only decrease 6 years after implementation, if at all (\cite{Agha2014TheCare}, \cite{dranove2014trillion}). More relevant to individual physicians' productivity, a number of studies consider the effect of EHRs on productivity in specific settings. One study finds that nursing home productivity increases after adopting health IT (\cite{Hitt2016}), but another finds that physician productivity decreases by 11 percent due to EMRs being adopted in primary-care sites (\cite{Meyerhoefer}).

I expand this literature, first by considering a different mechanism by which EHRs could indirectly affect patients through various physician responses, which, to my knowledge, has not yet been examined. This is the first empirical study that connects EHR implementation to physician labor market decisions, and focuses on the difference in response across age groups. Further, I improve on the health information technology literature by considering the primary time period in which EHRs were rapidly implemented. This provides sufficient variation in the timing of implementation and ensures the EHRs are advanced enough to be considered "meaningful" by the government's standards, and are likely affecting physicians' daily life. Similarly to past studies, I consider EHR implementation as a treatment variable in a difference-in-difference framework, but I improve on this by estimating average group time treatment effects for multiple years after implementation, which avoids common problems that arise in typical two way fixed effects estimation of heterogeneous effects, and further provides insight to short vs. long term effects taking place.

Early research on the inputs to physician retirement decisions found that the most influential factors are personal and financial matters, and that physicians care about their work environment (specifically in the context of managed care organizations), but do not necessarily retire early because of it (\cite{Bahrami2002}). Contrary to this work, I find that EHR implementation led to a 0.006 ppt (s.e. 0.002) increase in the likelihood of a senior physician retiring after implementation, a 20\% increase relative to the average retirement for this age group. Physicians less than 60 were 0.002 ppts (s.e. 0.0009) more likely to retire after exposure, possibly pointing to career switching since retirement is measured as a lack of Medicare patients seen. These results suggest that EHR implementation may have intensified access to care issues for patients, and that there was a surge of experienced physicians who left the labor force and were no longer available to influence physicians learning under them. For those who did not retire, I find no evidence of switching towards office settings in either age group. However, both age groups are more likely to change zip codes (by 0.045 ppts), suggesting across-hospital switching. Finally, I find a modest, positive effect on patient count (an increase of 14 patients relative to a mean of 330) completely driven by younger physicians, and no effect on billing activity for either age group. Thus, for younger physicians, EHRs allow the claim count to patient ratio to decrease, which, in the best case scenario, points to a decrease in wasteful care being billed for.

A key assumption that underlies the main analysis is that physicians are constrained to using the technology in a hospital who chooses to adopt. I consider two scenarios where this assumption would be violated. First, I investigate the existence of hospital employees with the purpose of using an EHR on behalf of a physician, which I refer to as data assistants. If data assistants are used to ease EHR use for physicians, the results are at least partially driven by their existence. I find that the majority of hiring of data assistants took place in 2013-2014, a two year delay from the vast EHR implementation. Further, my main results are not sensitive to the exclusion of these years, and estimating the effect of EHR implementation only in hospitals without data assistants yields noisy, but similar, results. Second, I investigate whether the decision for hospitals to implement EHRs is exogenous to individual physicians. That is, a physician does not choose their own EHR exposure. While this assumption is reasonable in many settings, there may be instances where a physician is involved in hospital decision making. For this reason, I also consider a sensitivity analysis in which I limit the sample only to hospitals where vertical integration between physician and the hospital is low, as this indicates a lower probability that EHR implementation is endogenous. The results of the additional analyses indicate that my main findings are not sensitive to endogeneity concerns stemming from joint decision making among physicians and hospitals regarding EHR adoptions.

@@

This section serves to outline the potential responses of physicians who work primarily in hospitals to EHRs based on the costs and incentives they face. The physician behaviors that I consider in this analysis are closely related and should not be taken as independent. For example, a physician's choice of work setting depends on their previous decision to prolong retirement, and billing activity depends on work setting. Thus, I discuss the behaviors in stages of decision making.

@@

Workers in developed countries tend to plan for formal retirement well in advance. Generally, an exogenous shock to a worker's environment is not expected to change retirement age due to the amount of wealth and planning necessary to formally leave the labor force. Classic retirement models structure the decision of retirement by maximizing expected lifetime utility based on future earnings and retirement benefits, or choosing to prolong retirement only if the expected earnings gain from doing so is positive (\cite{gustman1986disaggregated}, \cite{stock1990pension}). However, physicians typically make this decision differently than workers in other industries. Most physicians plan to retire at age 60, but do not actually retire until age 69 (\cite{collier2017challenges}). When the time to retire comes, many physicians hesitate to abandon patients they have seen over the course of their career. The decision to delay retirement is not financial, but altruistic. Thus, retiring may be in a physician's choice set years before the realized decision to leave the labor force.

It is important to note that age is the most important factor in the decision to retire, implying that a 35 year old physician has very different incentives than a 65 year old physician, and a technology shock will affect the age groups differently. First, I consider physicians already of retirement age when a new technology is implemented. While payment schemes, wealth, and retirement benefits certainly impact the utility a physician receives from working or retiring, I assume the decision to retire is independent of those things since, on average, those in this age group have already prepared financially to retire. Then, a physician will only retire if the utility from working is less than the utility from retiring, where utility from working depends on job satisfaction and personal life. Exogenous EHR implementation affects the decision to retire through its affect on job satisfaction. If EHRs impose a burden on daily tasks, there is an inverse relationship between EHR exposure and utility from continuing to work. If EHRs make daily tasks easier, there is a positive relationship. Thus, for a physician indifferent between retiring and continuing to work, EHRs induce retirement if they impose a cost in terms of job satisfaction, or induce continued working life if they impose a benefit to job satisfaction. Media articles focusing on interviews with particular physicians suggest that the implementation of EHRs did in fact lead to retirement in older physicians (\cite{ringel_2019}, \cite{loria_2020}), suggesting a negative impact.

Formal retirement (leaving the labor force altogether) is not the only way for physicians to transition out of a clinical role. There are opportunities for physicians of any age to switch careers towards more administrative roles such as teaching, consulting, or hospital management. While it is unlikely that physicians under 55 are formally retiring, physicians of any age could be career switching. This decision is impacted similarly to the above discussion regarding formal retirement, where a physician who is indifferent between clinical work and career switching can be affected one way or the other based on how EHRs impact job satisfaction of current clinical work. As I discuss in the Data Section, I only observe whether a physician stops seeing Medicare patients, not whether they retire or career switch. Thus, the term "retirement" for the purpose of this paper indicates no longer seeing patients, and makes no assertion about what the physician does afterwards. A physician is said to retire in a given year only if it is the first year in which the physician has no future Medicare claims.

@@

For most physicians, EHRs will not impose a costly enough burden to induce retirement. However, there are other ways physicians might change their behavior under sufficient incentive to avoid the technology. For those who remain practicing in a clinical environment, changing the physical place of work may allow the physician more control over the technologies used in the work environment, affecting overall job satisfaction. Generally, physicians may have the ability to move hospitals, move from a hospital to an office, or work in multiple facilities and shift the composition of patients towards other facilities. I only consider physicians who are classified as a primary care physician and are closely tied to at least one hospital, often referred to as hospitalists or internists (for a detailed discussion of included physicians, see Appendix \ref{app:data}). Some hospitalists work under contract full time for one medical practice, where they have a consistent shift schedule, while others contract themselves out to multiple hospitals, offices, or both.

First, I consider those who are in a relatively binding contract. For hospitalists under full time contract with a medical practice, the physician is more tethered to the hospital. This could impact the response to EHR implementation in that hospital in two ways: (1) through potentially increased use of the EHR in its initial phases, and (2) limited access to other work settings. A typical contract requires 90 to 120 days termination notice and some contracts include do not compete clauses (\cite{yasgur_by_-_yasgur_2016}), making a shift in work setting costly or excluding it from the choice set. Since these hospitalists are established in their hospital, they work a significant amount of time there. When an EHR is implemented, they may become familiar with it more quickly and be less likely to desire to switch work settings. Alternatively, there are hospitalists that work in multiple hospitals or work in both hospitals and offices. If one hospital implements a costly (or beneficial) EHR, these workers can more easily shift all or some of their work towards (or away from) different facilities to allocate to the place where they gain the most job satisfaction.

\subsection{Productivity}

There are likely still physicians who are not induced to change their behavior on the basis of retirement or work setting due to EHR implementation. A lack of response could be due to 3 factors: EHRs make no impact on the work environment, EHRs are costly, but not costly enough to shift, or EHRs improve job satisfaction in the current work environment. For physicians who make no physical changes to work environment after EHR exposure, it is natural to assess whether EHRs affect output in current work environment. An important, and heavily debated, aspect of the implementation of EHRs is whether or not they allow users to be more productive. A main purpose of the HITECH Act was to improve efficiency of care, specifically to decrease time spent on administrative burden. Such a decrease implies a change in patient care: either a larger number of patients have access to an appointment, or the same amount of patients receive more face-to-face time with physicians during appointments. Alternatively, because the technology is not hugely user friendly, it may not decrease administrative burden after all.

@@

The Medicare Data on Provider Practice and Specialty (MD-PPAS) is a database of physicians which details claim counts, physician specialty, and practice location from 2009-2017. Included in the MD-PPAS is physician NPI, primary and secondary specialties, both patient and total billing counts in up to 12 different zip codes, and fraction of patients in different settings such as office, inpatient or emergency room. I first limit the sample of physicians based on specialty type. My analysis relies on EHR implementation in hospitals. Thus, I only include physicians who reported their primary specialty as hospitalist (physician who self-identifies as hospitalist or has 90\% of patients in inpatient setting), internal medicine, general practice or family practice in at least one year. However, I exclude physicians who list themselves as a specialist in all but one year. Since the focus of the analysis is physicians working in hospitals, I also drop physicians with less than 70\% of patients in a hospital setting in all years and I drop physicians with insufficient claim counts to show indications of behavior.

@@

I show summary statistics for the entire sample in Table \ref{tab:sumstats}. Only 3\% of physicians (approximately 770) retire over the course of the panel. 30\% of the physician-year observations have value one for working in an office in some capacity, with about a tenth of the number of patients being seen in an office setting. By construction, all physicians are exposed to an EHR at some point in the sample, so the variation in treatment variables comes from the timing of exposure. Physicians are typically exposed about one fourth of the way through the sample, but there is a positive number of physicians first exposed in each year from 2009-2015. The average physician is 45 years old, works with 1.5 hospitals and 1.3 systems.

@@

I seek to estimate average treatment effects of EHR exposure on various physician outcomes. I treat EHR implementation as an exogenous treatment. In this setting, treatment effects are heterogeneous across physicians who are exposed in different years since the underlying characteristics of hospitals who adopt in different years may be correlated with the decision to adopt. Therefore, to avoid negative weighting issues this causes in a two way fixed effects specification, I estimate average treatment effects for a specific group $g$ at time $t$: 
$$ATT(g,t)=\mathbbm{E}[Y_t(g)-Y_t(0)|G_g=1],$$
where $G_g=1$ for those in group $g$. A group indicates all physicians treated, or first exposed to an EHR, in the same year. To estimate the heterogeneous treatment effects, I employ the doubly robust estimator established in \citeauthor{callaway2021difference} (\citeyear{callaway2021difference}). Other estimators that similarly address the concerns yield similar results, presented in Appendix \ref{app:estimators}. I also present classical two way fixed effects results with physician and year fixed effects in Appendix \ref{app:estimators}. These results yield similar patterns with slightly smaller magnitudes.

@@

 I consider the decision to switch work settings using multiple outcomes. First, I directly investigate the extent to which physicians work in office based setting and whether this changes due to EHR exposure. I use variation in two different variables: (1) an indicator variable equal to 1 if the physician has seen any positive number of patients in an office in a given year, and (2) the fraction of patients a physician sees in an office based setting in a given year. This analysis does not include physicians who retire at any point in the sample, as a zero could indicate dropping out of the data instead of working solely in a hospital.

@@

@@

\begin{figure}[p]
\centering
    \caption{Treatment Variables Over Time}
    \includegraphics[scale=.65]{Objects/sum_stats_year.pdf}
    \label{fig:treatmentgraph}
\end{figure}

@@

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Retirement}
    \includegraphics[scale=.65]{Objects/retire_plot.pdf}
    \label{fig:retirefirst}
\end{figure}

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Likelihood of Working in Office}
    \includegraphics[scale=.65]{Objects/officeind_plot.pdf}
    \label{fig:officefirst}
\end{figure}

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Fraction of Total Patients Seen in Office}
    \includegraphics[scale=.65]{Objects/officefrac_plot.pdf}
    \label{fig:officesecond}
\end{figure}

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Likelihood of Changing Zip Codes}
    \includegraphics[scale=.65]{Objects/zip_plot.pdf}
    \label{fig:zip}
\end{figure}

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Patient Count}
    \includegraphics[scale=.65]{Objects/patient_plot.pdf}
    \label{fig:patient}
\end{figure}

@@

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Claim Count, Limited Sample}
    \includegraphics[scale=.65]{Objects/claim_plot.pdf}
    \label{fig:claim}
\end{figure}

@@

I begin by connecting every National Provider Identifier (NPI)\footnote{Data can be downloaded from \hyperlink{https://download.cms.gov/nppes/NPI/Files.html}{\text{https://download.cms.gov/nppes/NPI-Files.html}}} to a description of their tax code\footnote{Data can be downloaded from \hyperlink{https://nucc.org/index.php/code-sets-mainmenu-41/provider-taxonomy-mainmenu-40/pdf-mainmenu-53}{\text{https://nucc.org/index.php/code-sets-mainmenu-41/provider-taxonomy-mainmenu-40/pdf-mainmenu-53}}}. I then categorize NPIs by key words in their tax code description. Any description containing ``Internal Medicine", ``Hospitalist", ``Family Medicine", or ``General Practice" is classified as a primary care physician (PCP), and any description containing ``hospital" is classified as a hospital. I create two data sets, one containing only PCPs and one containing only hospitals. These data are saved to use in identifying the relevant NPIs in the CMS Shared Patient Data.

@@

Before I present these estimators, I present results from a two way fixed effects setup that may suffer bias from negative weighting. These results from each outcome are presented in

@@

\begin{figure}
    \centering
    \includegraphics{objects}
    \caption{Caption}
    \label{fig:my_label}
\end{figure}
